---
title: "decoupleR_pathway_enrichment_analysis"
author: "Dustin Schilling"
date: "17 5 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Library 
```{r}
library(readr)
library(readxl)
library(cosmosR)
library(decoupleR)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(pheatmap)
library(patchwork)
library(cowplot)
```


# Input Data

```{r}
logFC_proteomics_z <- readRDS("results/logFC_proteomics_z.Rda")
logFC_metabolomics_z <- readRDS("results/logFC_metabolomics_z.Rda")

combined_metab_gene_hallmarks <- read_csv("support/combined_metab_gene_hallmarks.csv")

Biocrates_metabolite_identifier <- read_excel("support/Biocrates_metabolite_identifier.xlsx")
Biocrates_metabolite_identifier$feature <- gsub("[()/: -]",".",Biocrates_metabolite_identifier$feature)


network_combined <- combined_metab_gene_hallmarks
network_combined$mor <- 1
network_combined$term <- gsub("HALLMARK_","", network_combined$term)
network_combined$term <- gsub("_", " ", network_combined$term)

# network seperated
network_met <- network_combined[which(grepl("HMDB", network_combined$feature)),]
network_prot <- network_combined[!(network_combined$feature %in% network_met$feature),]

##################################
# limma ttest
##################################
proteomics_t <- readRDS("results/proteomics_DE_t.Rda")
metabolomics_t <- readRDS("results/metabolomics_DE_t.Rda")

proteomics_t <- proteomics_t[,c(3,6 )]
proteomics_t <- proteomics_t[-which(proteomics_t$ID == "NA"),]
proteomics_t<- unique(proteomics_t)
proteomics_t <- proteomics_t %>%
  group_by(ID) %>%
  summarise_each(funs(mean))
proteomics_t <- proteomics_t %>%
  drop_na() %>%
  column_to_rownames("ID") %>%
  as.matrix() 


metabolomics_t <- metabolomics_t[,c(12, 5)] 
metabolomics_t<- unique(metabolomics_t)
metabolomics_t <- metabolomics_t %>%
  group_by(HMDB) %>%
  summarise_each(funs(mean))
metabolomics_t <- metabolomics_t %>%
  drop_na() %>%
  column_to_rownames("HMDB") %>%
  as.matrix() 

combined_t <- rbind(metabolomics_t, proteomics_t)
```

# Data preparation
```{r}
#################################################
## metab
#################################################
metab_decoupler_z <- logFC_metabolomics_z %>%
  rownames_to_column(var= "ID")
metab_decoupler_z <- merge(metab_decoupler_z, Biocrates_metabolite_identifier, by.x = "ID", by.y = "feature")
metab_decoupler_z <- metab_decoupler_z[,-c(12:14)]

metab_decoupler_z <- metab_decoupler_z %>%
  mutate(HMDB = strsplit(as.character(HMDB), "/")) %>%
  unnest(HMDB) %>%
  filter(HMDB != "") %>%
  unique()

# average if metabolites have same HMDB number
# so there is only one HMDB as input
metab_decoupler_z <- metab_decoupler_z[,c(12,2:11)]
metab_decoupler_z <- metab_decoupler_z %>%
  group_by(HMDB) %>%
  summarise_each(funs(mean))
# now rownames can be annotated
metab_decoupler_z <- metab_decoupler_z %>%
  column_to_rownames("HMDB")
# NAs needs to be removed and input in matrix form
metab_decoupler_z <- metab_decoupler_z %>%
  drop_na() %>%
  as.matrix()
colnames(metab_decoupler_z) <- paste("Patient", colnames(metab_decoupler_z), sep ="_")


#################################################
## prot
#################################################
prot_decoupler_z <- logFC_proteomics_z %>%
  drop_na() %>%
  as.matrix()
colnames(prot_decoupler_z) <- paste("Patient", colnames(prot_decoupler_z), sep ="_")

#################################################
## combination
#################################################


combined_mat_z <- rbind(prot_decoupler_z, metab_decoupler_z)



##### Do the NA's cause problems -> drop_NA() commands deletes a lot of rows (HMDB, proteins)
##### Is there another approach to this?
```

# norm_wmean on DE genes - prot and metab seperated

```{r}
# Prot
res_wmean_PEA_prot <-  run_wmean(mat = proteomics_t, net = network_prot, .source = "term", .target = "feature", .mor="mor")
res_wmean_PEA_prot <- res_wmean_PEA_prot %>%
  filter(statistic == 'norm_wmean')

# Metab
res_wmean_PEA_met <-  run_wmean(mat = metabolomics_t, net = network_met, .source = "term", .target = "feature", .mor="mor")
res_wmean_PEA_met <- res_wmean_PEA_met %>%
  filter(statistic == 'norm_wmean')


pathway_select_prot <- res_wmean_PEA_prot %>%
  dplyr::filter(p_value < 0.05) %>%  # filter for significance
  dplyr::select("source", "score", "p_value") %>%
  dplyr::rename(pathway = "source", pvalue = "p_value") %>%
  dplyr::mutate(pathway = as.factor(pathway)) %>%
  dplyr::mutate(score = as.numeric(score)) %>%
  dplyr::arrange(pvalue) %>%
  slice(1:5) %>%
  dplyr::mutate(layer  = "proteomic")
pathway_select_met <- res_wmean_PEA_met %>%
  dplyr::filter(p_value < 0.05) %>%  # filter for significance
  dplyr::select("source", "score", "p_value") %>%
  dplyr::rename(pathway = "source", pvalue = "p_value") %>%
  dplyr::mutate(pathway = as.factor(pathway)) %>%
  dplyr::mutate(score = as.numeric(score)) %>%
  dplyr::arrange(pvalue) %>%
  slice(1:5) %>%
  dplyr::mutate(layer = "metabolomic")
pathway_select_sep <- rbind(pathway_select_met, pathway_select_prot)


```

# Visualize together
```{r}

colors1 <- c("steelblue", "indianred")
ggplot_norm_sep <-  ggplot(pathway_select_sep, aes(y = reorder(pathway, score), x = score, fill = layer)) +  
    geom_bar(stat = "identity") +
    scale_x_continuous(breaks = seq(-8, 12, by= 2)) +
    theme_bw() +
  geom_vline(xintercept = 0, linetype = 2) +
    theme(axis.title = element_text(face = "bold", size = 10),
          axis.text.y = element_text(size = 7),
          axis.text.x = element_text(size = 6)) +
    ylab("") +
  guides(fill=guide_legend(title="omic layer")) + 
    scale_fill_manual(values = colors1)


leg1 <- get_legend(ggplot_norm_sep)

p1 <- ggplot(pathway_select_prot, aes(y = reorder(pathway, score), x = score)) +
  geom_bar(stat = "identity", fill = colors[2])+
  theme_bw() + 
  geom_vline(xintercept = 0, linetype = 2) +
  theme(axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 6)) +
  ylab("")

m1 <- ggplot(pathway_select_met, aes(y = reorder(pathway, score), x = score)) +
  geom_bar(stat = "identity", fill =colors[1]) +
  theme_bw() +
  geom_vline(xintercept = 0, linetype = 2) +
  theme(axis.text.y = element_text(size = 7),
          axis.text.x = element_text(size = 6)) +
  ylab("")

comb_ggplot_sep <-  m1 + p1 


pm1 <- plot_grid(comb_ggplot_sep, leg1, rel_widths = c(3, .4))  + 
  plot_annotation(title = "Deregulated pathways in lung cancer (norm_wmean)",
  theme = theme(plot.title = element_text(size = 18)) & theme(plot.title = element_text(hjust = 0.5))
)
pm1

pdf("results/plots/norm_wmean_pathway_scores_HALLMARK_seperated.pdf", width = 12, height = 4)
pm1
dev.off()
```

# filter genes out of successful nodes
```{r}
# filter manually
dereg_pathway_met <- pathway_select_met$pathway
dereg_pathway_prot<- pathway_select_prot$pathway

deregulated_metabs <- data.frame(matrix(ncol = 2))
colnames(deregulated_metabs) <- c("feature", "term")
deregulated_genes <- data.frame(matrix(ncol = 2))
colnames(deregulated_genes) <- c("feature", "term")

for (i in 1:length(dereg_pathway_prot)) {
   x1 <- network_prot[which(network_prot$term %in% dereg_pathway_prot[i]), c(1,2) ]
   deregulated_genes <- rbind(deregulated_genes, x1)
}
deregulated_genes <- deregulated_genes[-1,]
deregulated_genes$layer <- "proteomic"
for (i in 1:length(dereg_pathway_met)) {
   x2 <- network_met[which(network_met$term %in% dereg_pathway_met[i]), c(1,2) ]
   deregulated_metabs <- rbind(deregulated_metabs, x2)
}
deregulated_metabs <- deregulated_metabs[-1,]
deregulated_metabs$layer <- "metabolomic"

deregulated_genes_and_metabs <- rbind(deregulated_genes, deregulated_metabs)

saveRDS(deregulated_genes_and_metabs, "results/deregulated_genes_and_metabs.RDS")
```


# norm_wmean on DE genes - combined
The combination of the the metabolic and proteomic layer in pathway enrichment analysis leads to completely different results, than on their own layer. Therefore the pathway of the 
```{r}
res_wmean_PEA_comb <-  run_wmean(mat = combined_t, net = network_combined, .source = "term", .target = "feature", .mor="mor")
res_wmean_PEA_comb <- res_wmean_PEA_comb %>%
  filter(statistic == 'norm_wmean')
```

# Visualization of all DE genes in pathways

```{r}
pathway_select_comb <- res_wmean_PEA_comb %>%
  dplyr::filter(p_value < 0.05) %>%  # filter for significance
  dplyr::select("source", "score", "p_value") %>%
  dplyr::rename(pathway = "source", pvalue = "p_value") %>%
  dplyr::mutate(pathway = as.factor(pathway)) %>%
  dplyr::mutate(score = as.numeric(score)) %>%
  dplyr::arrange(pvalue) %>%
  slice(1:10)

ggplot_normwmean <- ggplot(pathway_select_comb, aes(y = reorder(pathway, score), x = score)) +  
    geom_bar(aes(fill = score), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    scale_x_continuous(breaks = seq(-8, 12, by= 2)) +
    theme_bw() +
    theme(axis.title = element_text(face = "bold", size = 10),
          axis.text.y = element_text(size = 7),
          axis.text.x = element_text(size = 6)) +
    ylab("") +
    labs(title = "Normalized weighted mean scores in lung cancer \n for pathways in the Hallmarks gene set") + 
    theme(plot.title = element_text(hjust = 0.5))

pdf("results/plots/norm_wmean_pathway_scores_HALLMARK.pdf")
ggplot_normwmean
dev.off()
```

# norm_wmean on single patient level
```{r}
# take pathways from before

res_wmean_PEA_sp <-  run_wmean(mat = combined_mat_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
res_wmean_PEA_sp <- res_wmean_PEA_sp %>%
  filter(statistic == 'norm_wmean')

res_wmean_PEA_sp <- res_wmean_PEA_sp %>%
  select(condition, source, score) %>%
  dplyr::rename(pathway = "source") %>%
  dplyr::mutate(pathway = as.factor(pathway)) %>%
  dplyr::mutate(score = as.numeric(score)) %>%
  pivot_wider(names_from = condition, values_from = score)

subset.p <- which(res_wmean_PEA_sp$pathway %in% dereg_pathway)
  
res_wmean_PEA_sp <- res_wmean_PEA_sp[subset.p,]
res_wmean_PEA_sp <- res_wmean_PEA_sp %>%
  column_to_rownames("pathway")
```

# pheatmap for single patients
```{r}
p_res_PEA_sp <- pheatmap(res_wmean_PEA_sp)

pdf("results/plots/pheatmap_single_patients_decoupler_scores_norm_wmean.pdf")
p_res_PEA_sp
dev.off()
```

```{r}
sessionInfo()
```

