---
title: "decoupleR"
author: "Dustin Schilling"
date: "9 5 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(decoupleR)
library(readr)
library(readxl)
library(dplyr)
library(tibble)
library(tidyr)
```

## Input Data

```{r}
# network combined
combined_metab_gene_hallmarks <- read_csv("../support/combined_metab_gene_hallmarks.csv")
network_combined <- combined_metab_gene_hallmarks
network_combined$mor <- 1
network_combined$term <- gsub("HALLMARK_","", network_combined$term)
network_combined$term <- gsub("_", " ", network_combined$term)

# network separated
network_met <- network_combined[which(grepl("HMDB", network_combined$feature) == T),]
network_prot <- network_combined[which(grepl("HMDB", network_combined$feature) == F),]


# input for pathway enrichment analysis
logFC_PEA_prot_MTBE_z <- readRDS("../results/single_patient/logFC_PEA_prot_MTBE_z.Rda")
logFC_PEA_prot_autoSP3_z <- readRDS("../results/single_patient/logFC_PEA_prot_autoSP3_z.Rda")
logFC_PEA_metab_z <- readRDS("../results/single_patient/logFC_PEA_metab_z.Rda")


Biocrates_metabolite_identifier <- read_excel("../support/Biocrates_metabolite_identifier.xlsx")
Biocrates_metabolite_identifier$feature <- gsub("[()/: -]",".",Biocrates_metabolite_identifier$feature)
```

# Data transformation

```{r}
# transform metab from ID to HMDB numbers
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  rownames_to_column(var= "ID")
logFC_PEA_metab_z <- merge(logFC_PEA_metab_z, Biocrates_metabolite_identifier, by.x = "ID", by.y = "feature")
logFC_PEA_metab_z <- logFC_PEA_metab_z[,-c(6:8)]

logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  mutate(HMDB = strsplit(as.character(HMDB), "/")) %>%
  unnest(HMDB) %>%
  filter(HMDB != "") %>%
  unique()

# average if metabolites have same HMDB number
# so there is only one HMDB as input
logFC_PEA_metab_z <- logFC_PEA_metab_z[,c(6,2:5)]
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  group_by(HMDB) %>%
  summarise_each(funs(mean))
# now rownames can be annotated
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  column_to_rownames("HMDB")


# NAs needs to be removed and input in matrix form
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  drop_na() %>%
  as.matrix()
logFC_PEA_prot_MTBE_z <- logFC_PEA_prot_MTBE_z %>%
  drop_na() %>%
  as.matrix()
logFC_PEA_prot_autoSP3_z <- logFC_PEA_prot_autoSP3_z %>%
  drop_na() %>%
  as.matrix()

test <- logFC_PEA_metab_z
test[c(1:895), c(1:4)] <- as.numeric(1)
# now the inputs are all in the right format



## combined prot_MTBE with metab and prot_autoSP3 with metab
combined_MTBE <- rbind(logFC_PEA_metab_z, logFC_PEA_prot_MTBE_z)

combined_autoSP3 <- rbind(logFC_PEA_metab_z, logFC_PEA_prot_autoSP3_z)
```



# Single Pathway enrichment analysis with prot and metab combined!
Here, the metabolomic and proteomic data is combined. The pathway enrichment score are calculated then seperately for the extraction method MTBE and autoSP3. 
```{r}
# MTBE
norm_wmean_combined_MTBE <- run_wmean(mat = combined_MTBE, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_combined_MTBE <- norm_wmean_combined_MTBE %>%
  filter(statistic == 'norm_wmean')

# autoSP3
norm_wmean_combined_autoSP3 <- run_wmean(mat = combined_autoSP3, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_combined_autoSP3 <- norm_wmean_combined_autoSP3 %>%
  filter(statistic == 'norm_wmean')
```

# Correlation sample-wise
Compute correlation for the achieved scores in the pathway over each samples.
```{r}
# create matrix with pathway as rows and samples in columns
corr_matrix_MTBE <- norm_wmean_combined_MTBE[,c(3,2,4)]
corr_matrix_MTBE <- pivot_wider(corr_matrix_MTBE, names_from = condition, values_from = score )
corr_matrix_MTBE <- corr_matrix_MTBE %>%
  column_to_rownames("source")

corr_matrix_autoSP3 <- norm_wmean_combined_autoSP3[,c(3,2,4)]
corr_matrix_autoSP3 <- pivot_wider(corr_matrix_autoSP3, names_from = condition, values_from = score )
corr_matrix_autoSP3 <- corr_matrix_autoSP3 %>%
  column_to_rownames("source")

corr_samples <- diag(cor(corr_matrix_MTBE, corr_matrix_autoSP3, method= "pearson"))
## correlation sample wise
corr_samples

mean(corr_samples)
```

# Correlation over all samples -> do not consider
Compute correlation for the achieved scores in the pathway over all samples.
However this is is technically not correct, because then the mean in the correlation function would be taken over all samples.
```{r}
# correlation between the resulting scores of MTBE and autoSP3

cor(norm_wmean_combined_autoSP3$score, norm_wmean_combined_MTBE$score, method ="pearson")
```









# Single Pathway enrichment analysis seperated
```{r}
# metab 
norm_wmean_metab_sep <- run_wmean(mat = logFC_PEA_metab_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_metab_sep <- norm_wmean_metab_sep %>%
  filter(statistic == 'norm_wmean')

# prot
norm_wmean_prot_MTBE_sep <- run_wmean(mat = logFC_PEA_prot_MTBE_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_prot_MTBE_sep <- norm_wmean_prot_MTBE_sep %>%
  filter(statistic == 'norm_wmean')

norm_wmean_prot_autoSP3_sep <- run_wmean(mat = logFC_PEA_prot_autoSP3_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_prot_autoSP3_sep <- norm_wmean_prot_autoSP3_sep %>%
  filter(statistic == 'norm_wmean')

```







