---
title: "Differential_expression_analysis"
author: "Dustin Schilling"
date: "17 5 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Library

```{r}
library(cosmosR)
library(decoupleR)
library(dplyr)
library(tidyr)
library(pheatmap)
library(dendextend)
library(tidyverse)
library(ggrepel)
library(ggplot2)
library(metaboliteIDmapping)
library(readxl)
options(ggrepel.max.overlaps = Inf) 
library(patchwork)
library(stringr)
```

# Load decoupleR PEA and nodes of network
use prots and metabs of PEA and generated network and visualize them in volcano plot, PCA, pheatmap (logFC).

```{r}
deregulated_genes_and_metabs <- readRDS("results/deregulated_genes_and_metabs.RDS")
# Load succesfull nodes
nodes_lung_full<- readRDS("results/nodes_ORA_lung_full.RData")
nodes_lung<- readRDS("results/nodes_ORA_lung.RData")

dereg_PEA_network <- deregulated_genes_and_metabs[deregulated_genes_and_metabs$feature %in% nodes_lung_full$sucesses, ] 


# metab and prot
logFC_proteomics_z <- readRDS("results/logFC_proteomics_z.Rda")
logFC_metabolomics_z <- readRDS("results/logFC_metabolomics_z.Rda")

combined_metab_gene_hallmarks <- read_csv("support/combined_metab_gene_hallmarks.csv")

Biocrates_metabolite_identifier <- read_excel("support/Biocrates_metabolite_identifier.xlsx")
Biocrates_metabolite_identifier$feature <- gsub("[()/: -]",".",Biocrates_metabolite_identifier$feature)

##################################
# limma ttest
##################################
proteomics_t <- readRDS("results/proteomics_DE_t.Rda")
metabolomics_t <- readRDS("results/metabolomics_DE_t.Rda")

proteomics_t <- proteomics_t[,c(3,4, 6, 8 )]
proteomics_t <- proteomics_t[-which(proteomics_t$ID == "NA"),]
proteomics_t<- unique(proteomics_t)
proteomics_t <- proteomics_t %>%
  group_by(ID) %>%
  summarise_each(funs(mean))
proteomics_t <- proteomics_t %>%
  drop_na() %>%
  column_to_rownames("ID")


metabolomics_t <- metabolomics_t[,c(12, 3, 5, 7)] 
metabolomics_t<- unique(metabolomics_t)
metabolomics_t <- metabolomics_t %>%
  group_by(HMDB) %>%
  summarise_each(funs(mean))
metabolomics_t <- metabolomics_t %>%
  drop_na() %>%
  column_to_rownames("HMDB")

######################################################################
combined_t <- rbind(metabolomics_t, proteomics_t)
######################################################################

metabolomics_t <- metabolomics_t %>%
  rownames_to_column("HMDB")
metabolomics_t_map <- left_join(metabolomics_t, metabolitesMapping, by = "HMDB" )
metabolomics_t_map <-metabolomics_t_map %>%
  select(HMDB, logFC, t, adj.P.Val, Name) %>%
  unique()
metabolomics_t_map <- metabolomics_t_map[-which(is.na(metabolomics_t_map$Name) == T), ]

# only 4 HMDBs not annotated, these are:
metabolomics_t$HMDB[which(metabolomics_t$HMDB %in% metabolomics_t_map$HMDB == F)]

```

# Volcano Plot prot
```{r}
# format for filter 
volc.prot <- proteomics_t
p_cutoff = 0.001

volc.prot <- volc.prot %>%
  rename(p.val = "adj.P.Val") %>%
  mutate(Regulation =  ifelse(p.val <= p_cutoff & abs(logFC) >=2, ifelse(logFC>2, "up", "down"), "stable")) %>%
  rownames_to_column("ID")

volc.prot["significance"] <-  "not significant"
volc.prot[which(volc.prot$p.val < p_cutoff & abs(volc.prot$logFC) < 2 ),"significance"] <- "significant"
volc.prot[which(volc.prot$p.val > p_cutoff & abs(volc.prot$logFC) > 2 ),"significance"] <- "fold-change"
volc.prot[which(volc.prot$p.val < p_cutoff & abs(volc.prot$logFC) > 2 ),"significance"] <- "significant & fold-change"

volc.prot.filter <- volc.prot %>%
  filter(significance=="significant & fold-change")

prot.volc <-  ggplot(volc.prot, aes(x = logFC, y = -log10(p.val), col = Regulation, label = ID)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("blue", "grey","red")) +
  geom_hline(yintercept=-log10(p_cutoff), lty=4,col="black",lwd=0.8) +
  geom_vline(xintercept=c(-2,2),lty=4,col="black",lwd=0.8) +
   geom_text_repel(data=volc.prot.filter, aes(label=ID), show.legend = F) +
  ggtitle("Differentially Expressed Proteins in Lung Cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10 (adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5)
        #legend.title = element_blank(),
        #legend.text = element_text(size=10),
        #legend.key.size = unit(3,"line")
        )
prot.volc

pdf("results/plots/volcano_prot.pdf", width = 10)
prot.volc
dev.off()
```

# Volcano Plot metab
```{r}
# format for filter 
volc.met <- metabolomics_t_map
p_cutoff = 0.001


volc.met <- volc.met %>%
  rename(p.val = "adj.P.Val") %>%
  mutate(Regulation =  ifelse(p.val <= p_cutoff & abs(logFC) >=1, ifelse(logFC>1, "up", "down"), "stable"))

volc.met["significance"] <-  "not significant"
volc.met[which(volc.met$p.val < p_cutoff & abs(volc.met$logFC) < 1 ),"significance"] <- "significant"
volc.met[which(volc.met$p.val > p_cutoff & abs(volc.met$logFC) > 1 ),"significance"] <- "fold-change"
volc.met[which(volc.met$p.val < p_cutoff & abs(volc.met$logFC) > 1 ),"significance"] <- "significant & fold-change"

volc.met.filter <- volc.met %>%
  filter(significance=="significant & fold-change")

met.volc <-  ggplot(volc.met, aes(x = logFC, y = -log10(p.val), col = Regulation, label = Name)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("blue", "grey","red")) +
  geom_hline(yintercept=-log10(p_cutoff), lty=4,col="black",lwd=0.8) +
  geom_vline(xintercept=c(-1,1),lty=4,col="black",lwd=0.8) +
   geom_text_repel(data=volc.met.filter, aes(label=Name), show.legend = F) +
  ggtitle("Differentially Abundant Metabolites in Lung Cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10(adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), 
        #legend.position="right", 
        #legend.title = element_blank()
        )
met.volc

pdf("results/plots/volcano_metab.pdf", width = 10)
met.volc
dev.off()
```

# combine volcano plots
```{r}
combined <- prot.volc + met.volc & theme(legend.position = "right")
combined.volc <- combined + plot_layout(guides = "collect")
combined.volc

pdf("results/plots/volcano_combined.pdf", width = 12)
combined.volc +  plot_annotation(
  tag_levels = "A",
  tag_suffix = ')'
)
dev.off()
```



# PROTEINS AND METAB OUT OF NETWORK
e.g. deregulated_genes_and_metabs -> in volcano plot or heatmap, PCA
dereg_PEA_network, deregulated_genes_and_metabs, f_deregulated_genes_and_metabs
```{r}
volc.prot.x <- proteomics_t
p_cutoff = 0.001

volc.prot.x <- volc.prot.x %>%
  rename(p.val = "adj.P.Val") %>%
  rownames_to_column("ID")

volc.prot.x$network_appearance <- "not in network"
volc.prot.x$network_appearance[which(volc.prot.x$ID %in% nodes_lung_full[[1]])] <- "network"


volc.prot.x.filter <- volc.prot.x[which(volc.prot.x$ID %in% nodes_lung_full[[1]]), ]
volc.prot.x.filter

prot.volc.x <-  ggplot(volc.prot.x, aes(x = logFC, y = -log10(p.val), col = network_appearance, label = ID)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("red","grey")) + 
  geom_hline(yintercept=-log10(p_cutoff), lty=4,col="black",lwd=0.8) +
  geom_vline(xintercept=c(-2,2),lty=4,col="black",lwd=0.8) +
   geom_text_repel(data=volc.prot.x.filter, aes(label=ID), show.legend = F) +
  ggtitle("Nodes of COSMOS network in Lung Cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10 (adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5)
        #legend.title = element_blank(),
        #legend.text = element_text(size=10),
        #legend.key.size = unit(3,"line")
        )
prot.volc.x
```

# MYC TARGETS PEA in volcano plots metab and prot
```{r}
MYC_deregulated_genes_and_metabs <- deregulated_genes_and_metabs %>%
  filter(term == "MYC TARGETS V1")

volc.prot.MYC <- proteomics_t
p_cutoff = 0.01

volc.prot.MYC <- volc.prot.MYC %>%
  rename(p.val = "adj.P.Val") %>%
  rownames_to_column("ID")

volc.prot.MYC$MYC_targets <- "no target of MYC"
volc.prot.MYC$MYC_targets[which(volc.prot.MYC$ID %in% MYC_deregulated_genes_and_metabs$feature)] <- "MYC target"
  
##
volc.prot.MYC.filter <- volc.prot.MYC[which(volc.prot.MYC$ID %in% MYC_deregulated_genes_and_metabs$feature), ]
#filter for labeling significant Proteins in MYC pathway
volc.prot.MYC.filter <- volc.prot.MYC.filter %>%
  filter(p.val < p_cutoff & abs(logFC) >= 1)
##


prot.volc.MYC <-  ggplot(volc.prot.MYC, aes(x = logFC, y = -log10(p.val), col = MYC_targets, label = ID)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("red","grey")) + 
  geom_hline(yintercept=-log10(p_cutoff), lty=4,col="black",lwd=0.8) +
  geom_vline(xintercept=c(-1,1),lty=4,col="black",lwd=0.8) +
  geom_text_repel(data=volc.prot.MYC.filter, aes(label=ID), show.legend = F) +
  ggtitle("MYC Targets in Proteomics of Lung Cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10 (adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5)
        #legend.title = element_blank(),
        #legend.text = element_text(size=10),
        #legend.key.size = unit(3,"line")
        )
prot.volc.MYC


############################################################################
############################################################################


MYC_deregulated_genes_and_metabs <- deregulated_genes_and_metabs %>%
  filter(term == "MYC TARGETS V1")

volc.met.MYC <- metabolomics_t_map
p_cutoff = 0.01

volc.met.MYC <- volc.met.MYC %>%
  rename(p.val = "adj.P.Val")

volc.met.MYC$MYC_targets <- "no target of MYC"
volc.met.MYC$MYC_targets[which(volc.met.MYC$HMDB %in% MYC_deregulated_genes_and_metabs$feature)] <- "MYC target"
  
###
volc.met.MYC.filter <- volc.met.MYC[which(volc.met.MYC$HMDB %in% MYC_deregulated_genes_and_metabs$feature), ]
#filter for labeling significant metabolites in MYC pathway


#1,4−Butanediamine, N,N'−bis(3−aminopropyl)− is spermine
volc.met.MYC.filter[8, "Name"] <- "Spermine"

##


met.volc.MYC <-  ggplot(volc.met.MYC, aes(x = logFC, y = -log10(p.val), col = MYC_targets, label = Name)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("red","grey")) +
  geom_text_repel(data=volc.met.MYC.filter, aes(label=Name), show.legend = F) +
  ggtitle("MYC targets in Metabolomics of Lung Cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10 (adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5)
        #legend.title = element_blank(),
        #legend.text = element_text(size=10),
        #legend.key.size = unit(3,"line")
        )
met.volc.MYC



combined_MYC <-  prot.volc.MYC + met.volc.MYC & theme(legend.position = "right")
combined.volc.MYC <- combined_MYC + plot_layout(guides = "collect")
combined.volc.MYC


pdf("results/plots/MYC_targets_volcano_plot.pdf", width = 12)
combined.volc.MYC +  plot_annotation(
  tag_levels = "A",
   tag_suffix = ')'
)
dev.off()


```




```{r}
sessionInfo()
```

