---
title: "Differential_expression_analysis"
author: "Dustin Schilling"
date: "17 5 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Library

```{r}
library(cosmosR)
library(decoupleR)
library(dplyr)
library(tidyr)
library(pheatmap)
library(dendextend)
library(tidyverse)
library(ggrepel)
library(ggplot2)
library(readxl)
options(ggrepel.max.overlaps = Inf) 
library(patchwork)
library(stringr)
```

# Load decoupleR PEA and nodes of network
Use prots and metabs of PEA and generated network and visualize them in volcano plot (logFC).
deregulated_genes_and_metabs is 
```{r}
deregulated_genes_and_metabs <- readRDS("results/deregulated_genes_and_metabs.RDS")
deregulated_prot <- deregulated_genes_and_metabs %>%
  filter(layer == "proteomic")
deregulated_met <-  deregulated_genes_and_metabs %>%
  filter(layer == "metabolomic")

# Load succesfull nodes
nodes_lung_full<- readRDS("results/nodes_ORA_lung_full.RData")
nodes_lung<- readRDS("results/nodes_ORA_lung.RData")

# metab and prot
logFC_proteomics_z <- readRDS("results/logFC_proteomics_z.Rda")
logFC_metabolomics_z <- readRDS("results/logFC_metabolomics_z.Rda")

Biocrates_metabolite_identifier <- read_excel("support/Biocrates_metabolite_identifier.xlsx")
Biocrates_metabolite_identifier$feature <- gsub("[()/: -]",".",Biocrates_metabolite_identifier$feature)

##################################
# limma ttest
##################################
metabolomics_DE_t <- as.data.frame(read_delim("data/metabolomics_DE_t_TUvsNG.txt", 
                                              delim = "\t", escape_double = FALSE, 
                                              trim_ws = TRUE))

metabolomics_DE_t <- merge(metabolomics_DE_t, Biocrates_metabolite_identifier, by.x = "ID", by.y = "feature")
metabolomics_DE_t <- metabolomics_DE_t %>%
  mutate(HMDB = strsplit(as.character(HMDB), "/")) %>%
  unnest(HMDB) %>%
  filter(HMDB != "")
metabolomics_t <- metabolomics_DE_t
metabolomics_DE_t <- metabolomics_DE_t[,c(1, 3, 5, 7, 12)] 
metabolomics_t <- metabolomics_t[,c(1, 3, 5, 7, 12)] 
multipleID <- unique(metabolomics_DE_t$ID)
metabolomics_t <- metabolomics_t[0, ]
for (i in 1:length(multipleID)) {
  x <- which(metabolomics_DE_t$ID %in% multipleID[i])[1]
  metabolomics_t <- rbind(metabolomics_t, metabolomics_DE_t[x,])
}
# use metabolics_t
####################################

proteomics_t <- readRDS("results/proteomics_DE_t.Rda")

proteomics_t <- proteomics_t[,c(3,4, 6, 8 )]
proteomics_t <- proteomics_t[-which(proteomics_t$ID == "NA"),]
multipleID_prot <- unique(proteomics_t$ID)
proteomics_DE_t <- proteomics_t
proteomics_t <- proteomics_t[0,]
for (i in 1:length(multipleID_prot)){
  x1 <- which(proteomics_DE_t$ID %in% multipleID_prot[i])[1]
  proteomics_t <- rbind(proteomics_t, proteomics_DE_t[x1,])
}
proteomics_t <- proteomics_t[!is.na(proteomics_t$logFC),]

# use proteomics_t
####################################

```

# Volcano Plot prot
```{r}
# format for filter 
volc.prot <- proteomics_t
p_cutoff = 0.001

volc.prot <- volc.prot %>%
  rename(p.val = "adj.P.Val") %>%
  mutate(Regulation =  ifelse(p.val <= p_cutoff & abs(logFC) >=2, ifelse(logFC>2, "up", "down"), "stable"))

volc.prot["significance"] <-  "not significant"
volc.prot[which(volc.prot$p.val < p_cutoff & abs(volc.prot$logFC) < 2 ),"significance"] <- "significant"
volc.prot[which(volc.prot$p.val > p_cutoff & abs(volc.prot$logFC) > 2 ),"significance"] <- "fold-change"
volc.prot[which(volc.prot$p.val < p_cutoff & abs(volc.prot$logFC) > 2 ),"significance"] <- "significant & fold-change"

volc.prot.filter <- volc.prot %>%
  filter(significance=="significant & fold-change")

prot.volc <-  ggplot(volc.prot, aes(x = logFC, y = -log10(p.val), col = Regulation, label = ID)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("blue", "grey","red")) +
  geom_hline(yintercept=-log10(p_cutoff), lty=4,col="black",lwd=0.8) +
  geom_vline(xintercept=c(-2,2),lty=4,col="black",lwd=0.8) +
   geom_text_repel(data=volc.prot.filter, aes(label=ID), show.legend = F) +
  ggtitle("Differentially expressed proteins in lung cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10(adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5)
        #legend.title = element_blank(),
        #legend.text = element_text(size=10),
        #legend.key.size = unit(3,"line")
        )
prot.volc

pdf("results/plots/volcano_prot.pdf", width = 10)
prot.volc
dev.off()
```

# Volcano Plot metab
```{r}
# format for filter 
volc.met <- metabolomics_t
p_cutoff = 0.001


volc.met <- volc.met %>%
  rename(p.val = "adj.P.Val") %>%
  mutate(Regulation =  ifelse(p.val <= p_cutoff & abs(logFC) >=1, ifelse(logFC>1, "up", "down"), "stable"))

volc.met["significance"] <-  "not significant"
volc.met[which(volc.met$p.val < p_cutoff & abs(volc.met$logFC) < 1 ),"significance"] <- "significant"
volc.met[which(volc.met$p.val > p_cutoff & abs(volc.met$logFC) > 1 ),"significance"] <- "fold-change"
volc.met[which(volc.met$p.val < p_cutoff & abs(volc.met$logFC) > 1 ),"significance"] <- "significant & fold-change"

volc.met.filter <- volc.met %>%
  filter(significance=="significant & fold-change")

met.volc <-  ggplot(volc.met, aes(x = logFC, y = -log10(p.val), col = Regulation, label = ID)) +
  geom_point(alpha=0.4, size=2) +
  scale_color_manual(values=c("blue", "grey","red")) +
  geom_hline(yintercept=-log10(p_cutoff), lty=4,col="black",lwd=0.8) +
  geom_vline(xintercept=c(-1,1),lty=4,col="black",lwd=0.8) +
   geom_text_repel(data=volc.met.filter, aes(label=ID), show.legend = F) +
  ggtitle("Differentially abundant metabolites in lung cancer") +
  xlab("log2(fold change)") + 
  ylab("-log10(adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), 
        #legend.position="right", 
        #legend.title = element_blank()
        )
met.volc

pdf("results/plots/volcano_metab.pdf", width = 10)
met.volc
dev.off()
```

# combine volcano plots
```{r}
combined <- prot.volc + met.volc & theme(legend.position = "right")
combined.volc <- combined + plot_layout(guides = "collect")
combined.volc

pdf("results/plots/volcano_combined.pdf", width = 12)
combined.volc +  plot_annotation(
  tag_levels = "A",
  tag_suffix = ')'
)
dev.off()
```

# main drivers in pathways of PEA
# met
Select of every pathway the top 5 main drivers -> highest abs t- value
```{r}
# deregulated_prot
# deregulated_met
Biocrates_metabolite_identifier <- Biocrates_metabolite_identifier %>%
  mutate(HMDB = strsplit(as.character(HMDB), "/")) %>%
  unnest(HMDB) %>%
  filter(HMDB != "")
deregulated_met <- deregulated_met %>%
  dplyr::rename(mapping = "feature")
dereg_metabolomics <- merge(deregulated_met, Biocrates_metabolite_identifier, by.x = "mapping", by.y = "HMDB")
dereg_metabolomics <- dereg_metabolomics[,c(1,2,4)]
dereg_metabolomics_t <- merge(metabolomics_t, dereg_metabolomics, by.x = "ID", by.y ="feature")
dereg_metabolomics_t <- dereg_metabolomics_t[,-6]
dereg_metabolomics_t <- dereg_metabolomics_t[!duplicated(dereg_metabolomics_t),]

# Select of every pathway the top 5 main drivers -> highest abs t- value
dereg_metabolomics_t <- dereg_metabolomics_t %>%
  group_by(term) %>%
  arrange(desc(abs(t))) %>%
  slice(1:5)
```

# Volc plot met
```{r}
# volcano plot
volc.d.met <- metabolomics_t
volc.d.met <- left_join(volc.d.met, dereg_metabolomics_t)
subset.missing <- which(is.na(volc.d.met$term))
volc.d.met[subset.missing,6] <- "not relevant"
colnames(volc.d.met)[6] <- "pathway"
multiplevolcID <- unique(dereg_metabolomics_t$ID)
volc.d.met.filter <- dereg_metabolomics_t
volc.d.met.filter <- volc.d.met.filter[0, ]
for (i in 1:length(multiplevolcID)) {
  x <- which(dereg_metabolomics_t$ID %in% multiplevolcID[i])[1]
  volc.d.met.filter <- rbind(volc.d.met.filter, dereg_metabolomics_t[x,])
}
# volc.d.met.filter -> for labelling


volc.PEA.drivers <-  ggplot(volc.d.met, aes(x = logFC, y = -log10(adj.P.Val), col = pathway, alpha=factor(pathway))) +
  geom_point(size=3,  position = position_jitter(w =0.04)) +
  geom_text_repel(data=volc.d.met.filter, aes(x = logFC, y = -log10(adj.P.Val), label=ID), show.legend = F, inherit.aes = F) +
  scale_color_manual(values = c("ANDROGEN RESPONSE" = "blue", "APICAL JUNCTION" = "red", "G2M CHECKPOINT" = "orange", "GLYCOLYSIS" = "green", "HYPOXIA" = "purple", "not relevant" = "grey")) +
  scale_alpha_manual(values = c("ANDROGEN RESPONSE" = 0.8, "APICAL JUNCTION" = 0.8, "G2M CHECKPOINT" = 0.8, "GLYCOLYSIS" = 0.8, "HYPOXIA" = 0.8, "not relevant" = 0.4), guide = "none") +
  ggtitle("Main drivers of deregulated pathways in metabolomics") +
  xlab("log2(fold change)") + 
  ylab("-log10(adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), 
        #legend.position="right", 
        #legend.title = element_blank()
        ) 
volc.PEA.drivers

pdf("results/plots/main_drivers_PEA_metab.pdf", width = 10)
volc.PEA.drivers
dev.off()
```

# prot
Select of every pathway the top 5 main drivers -> highest abs t- value
```{r}
# deregulated_prot
# deregulated_met
dereg_proteomics_t <- merge(proteomics_t, deregulated_prot, by.x = "ID", by.y ="feature")
dereg_proteomics_t <- dereg_proteomics_t[,-6]
dereg_proteomics_t <- dereg_proteomics_t[!duplicated(dereg_proteomics_t),]

# Select of every pathway the top 5 main drivers -> highest abs t- value
dereg_proteomics_t <- dereg_proteomics_t %>%
  group_by(term) %>%
  arrange(desc(abs(t))) %>%
  slice(1:5)
```

# Volc plot met
```{r}
# volcano plot
volc.d.prot <- proteomics_t
volc.d.prot <- left_join(volc.d.prot, dereg_proteomics_t)
subset.missing <- which(is.na(volc.d.prot$term))
volc.d.prot[subset.missing,5] <- "not relevant"
colnames(volc.d.prot)[5] <- "pathway"
multiplevolcID_prot <- unique(dereg_proteomics_t$ID)
volc.d.prot.filter <- dereg_proteomics_t
volc.d.prot.filter <- volc.d.prot.filter[0, ]
for (i in 1:length(multiplevolcID_prot)) {
  x <- which(dereg_proteomics_t$ID %in% multiplevolcID_prot[i])[1]
  volc.d.prot.filter <- rbind(volc.d.prot.filter, dereg_proteomics_t[x,])
}
# volc.d.prot.filter -> for labelling


volc.PEA.drivers.prot <-  ggplot(volc.d.prot, aes(x = logFC, y = -log10(adj.P.Val), col = pathway, alpha=factor(pathway))) +
  geom_point(size=3,  position = position_jitter(w =0.04)) +
  geom_text_repel(data=volc.d.prot.filter, aes(x = logFC, y = -log10(adj.P.Val), label=ID), show.legend = F, inherit.aes = F) +
  scale_color_manual(values = c("COAGULATION" = "blue", "E2F TARGETS" = "red", "EPITHELIAL MESENCHYMAL TRANSITION" = "orange", "G2M CHECKPOINT" = "green", "GLYCOLYSIS" = "purple", "not relevant" = "grey")) +
  scale_alpha_manual(values = c("COAGULATION" = 0.8, "E2F TARGETS" = 0.8, "EPITHELIAL MESENCHYMAL TRANSITION" = 0.8, "G2M CHECKPOINT" = 0.8, "GLYCOLYSIS" = 0.8, "not relevant" = 0.4), guide = "none") +
  ggtitle("Main drivers of deregulated pathways in proteomics") +
  xlab("log2(fold change)") + 
  ylab("-log10(adj. p-value)") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), 
        #legend.position="right", 
        #legend.title = element_blank()
        ) 
volc.PEA.drivers.prot

pdf("results/plots/main_drivers_PEA_proteomics.pdf", width = 10)
volc.PEA.drivers.prot
dev.off()
```



```{r}
sessionInfo()
```

