---
title: "decoupleR"
author: "Dustin Schilling"
date: "9 5 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(decoupleR)
library(readr)
library(readxl)
library(dplyr)
library(tibble)
library(tidyr)
library(pheatmap)
library(ggplot2)
library(ComplexHeatmap)

```

## Input Data

```{r}
# network combined

combined_metab_gene_hallmarks <- read_csv("support/combined_metab_gene_hallmarks.csv")
network_combined <- combined_metab_gene_hallmarks
network_combined$mor <- 1
network_combined$term <- gsub("HALLMARK_","", network_combined$term)
network_combined$term <- gsub("_", " ", network_combined$term)

# network separated
network_met <- network_combined[which(grepl("HMDB", network_combined$feature) == T),]
network_prot <- network_combined[which(grepl("HMDB", network_combined$feature) == F),]


# input for pathway enrichment analysis
logFC_PEA_prot_MTBE_z <- readRDS("results/single_patient/logFC_PEA_prot_MTBE_z.Rda")
logFC_PEA_prot_autoSP3_z <- readRDS("results/single_patient/logFC_PEA_prot_autoSP3_z.Rda")
logFC_PEA_metab_z <- readRDS("results/single_patient/logFC_PEA_metab_z.Rda")


Biocrates_metabolite_identifier <- read_excel("support/Biocrates_metabolite_identifier.xlsx")
Biocrates_metabolite_identifier$feature <- gsub("[()/: -]",".",Biocrates_metabolite_identifier$feature)
```

# Data transformation

```{r}
# transform metab from ID to HMDB numbers
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  rownames_to_column(var= "ID")
logFC_PEA_metab_z <- merge(logFC_PEA_metab_z, Biocrates_metabolite_identifier, by.x = "ID", by.y = "feature")
logFC_PEA_metab_z <- logFC_PEA_metab_z[,-c(6:8)]

logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  mutate(HMDB = strsplit(as.character(HMDB), "/")) %>%
  unnest(HMDB) %>%
  filter(HMDB != "") %>%
  unique()

# average if metabolites have same HMDB number
# so there is only one HMDB as input
logFC_PEA_metab_z <- logFC_PEA_metab_z[,c(6,2:5)]
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  group_by(HMDB) %>%
  summarise_each(funs(mean))
# now rownames can be annotated
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  column_to_rownames("HMDB")


# NAs needs to be removed and input in matrix form
logFC_PEA_metab_z <- logFC_PEA_metab_z %>%
  drop_na() %>%
  as.matrix()
colnames(logFC_PEA_metab_z) <- paste("Patient", colnames(logFC_PEA_metab_z), sep ="_")
logFC_PEA_prot_MTBE_z <- logFC_PEA_prot_MTBE_z %>%
  drop_na() %>%
  as.matrix()
colnames(logFC_PEA_prot_MTBE_z) <- paste("Patient", colnames(logFC_PEA_prot_MTBE_z), sep ="_")
logFC_PEA_prot_autoSP3_z <- logFC_PEA_prot_autoSP3_z %>%
  drop_na() %>%
  as.matrix()
colnames(logFC_PEA_prot_autoSP3_z) <- paste("Patient", colnames(logFC_PEA_prot_autoSP3_z), sep ="_")

# now the inputs are all in the right format



## combined prot_MTBE with metab and prot_autoSP3 with metab
combined_MTBE <- rbind(logFC_PEA_metab_z, logFC_PEA_prot_MTBE_z)

combined_autoSP3 <- rbind(logFC_PEA_metab_z, logFC_PEA_prot_autoSP3_z)
```


# Single Pathway enrichment analysis with prot and metab combined!
Here, the metabolomic and proteomic data is combined. The pathway enrichment score are calculated then seperately for the extraction method MTBE and autoSP3. 

Integrated Metabolomics
```{r}
# MTBE
norm_wmean_combined_MTBE <- run_wmean(mat = combined_MTBE, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_combined_MTBE <- norm_wmean_combined_MTBE %>%
  filter(statistic == 'norm_wmean')

# autoSP3
norm_wmean_combined_autoSP3 <- run_wmean(mat = combined_autoSP3, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_combined_autoSP3 <- norm_wmean_combined_autoSP3 %>%
  filter(statistic == 'norm_wmean')
```

# Correlation sample-wise
Compute correlation for the achieved scores in the pathway over each samples.

corr: prot1_metab -> prot2_metab
```{r}
# create matrix with pathway as rows and samples in columns
corr_matrix_MTBE <- norm_wmean_combined_MTBE[,c(3,2,4)]
corr_matrix_MTBE <- pivot_wider(corr_matrix_MTBE, names_from = condition, values_from = score )
corr_matrix_MTBE <- corr_matrix_MTBE %>%
  column_to_rownames("source")

corr_matrix_autoSP3 <- norm_wmean_combined_autoSP3[,c(3,2,4)]
corr_matrix_autoSP3 <- pivot_wider(corr_matrix_autoSP3, names_from = condition, values_from = score )
corr_matrix_autoSP3 <- corr_matrix_autoSP3 %>%
  column_to_rownames("source")

corr_integrated <- diag(cor(corr_matrix_MTBE, corr_matrix_autoSP3, method= "pearson"))
## correlation sample wise
corr_integrated

mean_corr_integrated <- mean(corr_integrated)
mean_corr_integrated

```

# only prot1 and prot2 PEA correlation
Proteomics

```{r}
# metab 
norm_wmean_metab_sep <- run_wmean(mat = logFC_PEA_metab_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_metab_sep <- norm_wmean_metab_sep %>%
  filter(statistic == 'norm_wmean')

# prot
norm_wmean_prot_MTBE_sep <- run_wmean(mat = logFC_PEA_prot_MTBE_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_prot_MTBE_sep <- norm_wmean_prot_MTBE_sep %>%
  filter(statistic == 'norm_wmean')

norm_wmean_prot_autoSP3_sep <- run_wmean(mat = logFC_PEA_prot_autoSP3_z, net = network_combined, .source = "term", .target = "feature", .mor="mor")
norm_wmean_prot_autoSP3_sep <- norm_wmean_prot_autoSP3_sep %>%
  filter(statistic == 'norm_wmean')

```



```{r}
corr_matrix_MTBE_sep <- norm_wmean_prot_MTBE_sep[,c(3,2,4)]
corr_matrix_MTBE_sep <- pivot_wider(corr_matrix_MTBE_sep, names_from = condition, values_from = score )
corr_matrix_MTBE_sep <- corr_matrix_MTBE_sep %>%
  column_to_rownames("source")

corr_matrix_autoSP3_sep <- norm_wmean_prot_autoSP3_sep[,c(3,2,4)]
corr_matrix_autoSP3_sep <- pivot_wider(corr_matrix_autoSP3_sep, names_from = condition, values_from = score )
corr_matrix_autoSP3_sep <- corr_matrix_autoSP3_sep %>%
  column_to_rownames("source")


corr_proteomics <- diag(cor(corr_matrix_MTBE_sep, corr_matrix_autoSP3_sep, method= "pearson"))
## correlation sample wise
corr_proteomics 
 
mean_corr_proteomics <- mean(corr_proteomics)
mean_corr_proteomics
```

```{r}
simplified_corr_matrix_MTBE_sep <- corr_matrix_MTBE_sep
simplified_corr_matrix_MTBE_sep[abs(simplified_corr_matrix_MTBE_sep) < 2.5] <- 0
simplified_corr_matrix_MTBE_sep <- sign(simplified_corr_matrix_MTBE_sep)

simplified_corr_matrix_autoSP3_sep <- corr_matrix_autoSP3_sep
simplified_corr_matrix_autoSP3_sep[abs(simplified_corr_matrix_autoSP3_sep) < 2.5] <- 0
simplified_corr_matrix_autoSP3_sep <- sign(simplified_corr_matrix_autoSP3_sep)



difference_proteo_hallmarks <- as.matrix(simplified_corr_matrix_MTBE_sep - simplified_corr_matrix_autoSP3_sep)
colnames(difference_proteo_hallmarks) <- c("Patient_A","Patient_B","Patient_C","Patient_D")

heatmap_plot <- ComplexHeatmap::pheatmap(abs(difference_proteo_hallmarks[rowMaxs(abs(difference_proteo_hallmarks)) > 0,]), display_numbers = F, cluster_cols = F,  cluster_rows=T, fontsize_number = 14, treeheight_row = 0, color = cols, breaks = breaks, column_names_side = c("top"), fontsize = 12, angle_col = c("0"), annotation_legend =T)
heatmap_plot

heatmap_plot <- ComplexHeatmap::pheatmap(abs(difference_proteo_hallmarks), display_numbers = F, cluster_cols = F,  cluster_rows=T, fontsize_number = 14, treeheight_row = 0, color = cols, breaks = breaks, column_names_side = c("top"), fontsize = 12, angle_col = c("0"), annotation_legend =T)
heatmap_plot
```
# proteomic integrated average

Integrated average

```{r}
matrix_metab <- norm_wmean_metab_sep[,c(3,2,4)]
matrix_metab <- pivot_wider(matrix_metab, names_from = condition, values_from = score )


matrix_MTBE <- norm_wmean_prot_MTBE_sep[,c(3,2,4)]
matrix_MTBE <- pivot_wider(matrix_MTBE, names_from = condition, values_from = score )

matrix_autoSP3 <- norm_wmean_prot_autoSP3_sep[,c(3,2,4)]
matrix_autoSP3 <- pivot_wider(matrix_autoSP3, names_from = condition, values_from = score )


# filtering of same pathways between metab and MTBE, autoSP3 respectively

subset_corr <- which(matrix_metab$source %in% matrix_MTBE$source)
source_subset <- matrix_metab$source[subset_corr]

matrix_metab <- matrix_metab[subset_corr,]
matrix_metab <- matrix_metab %>%
  column_to_rownames("source")


matrix_MTBE <- matrix_MTBE[which(matrix_MTBE$source %in% source_subset),]
matrix_MTBE <- matrix_MTBE %>%
  column_to_rownames("source")


matrix_autoSP3 <- matrix_autoSP3[which(matrix_autoSP3$source %in% source_subset),]
matrix_autoSP3 <- matrix_autoSP3 %>%
  column_to_rownames("source")
```

```{r}
# average metab and proteomics (prot1, prot2 respectively)
corr_matrix_MTBE_metab <- list(matrix_metab, matrix_MTBE)
corr_matrix_MTBE_metab_avg <- Reduce("+",corr_matrix_MTBE_metab)/length(corr_matrix_MTBE_metab)

corr_matrix_autoSP3_metab <- list(matrix_metab, matrix_autoSP3)
corr_matrix_autoSP3_metab_avg <- Reduce("+",corr_matrix_autoSP3_metab)/length(corr_matrix_autoSP3_metab)
```

```{r}
corr_integrated_average <- diag(cor(corr_matrix_MTBE_metab_avg, corr_matrix_autoSP3_metab_avg, method ="pearson"))
corr_integrated_average
mean_corr_integrated_average<- mean(corr_integrated_average)
mean_corr_integrated_average
```




# Visualize in table
```{r}
ph_corr <- matrix(NA, ncol = 3, nrow=5)

colnames(ph_corr) <- c("Proteomics", "Integrated", "Integrated Average")
rownames(ph_corr) <- c("Patient_A", "Patient_B", "Patient_C", "Patient_D", "Average")


ph_corr[c(1:4),1] <- corr_proteomics
ph_corr[5,1] <- mean_corr_proteomics

ph_corr[c(1:4),2] <- corr_integrated
ph_corr[5,2] <- mean_corr_integrated

ph_corr[c(1:4),3] <- corr_integrated_average
ph_corr[5,3] <- mean_corr_integrated_average

breaks <- c(seq(0,0.99, by=.01))
cols <- c(colorRampPalette(c("white","#7285A5"))(length(breaks))) # or use length(breaks)


heatmap_plot <- ComplexHeatmap::pheatmap(ph_corr, display_numbers = T, cluster_cols = F,  cluster_rows=F, fontsize_number = 14, treeheight_row = 0, color = cols, breaks = breaks, column_names_side = c("top"), fontsize = 12, angle_col = c("0"), annotation_legend =T)
heatmap_plot


pdf("pHeatmap_pathway_enrichment_scores_corr_proteomics_color.pdf", height = 3.5)
heatmap_plot
dev.off()

```

# corr between metab and prot1/2
```{r}
# metab and MTBE
corr_metab_MTBE <- diag(cor(matrix_metab, matrix_MTBE, method= "pearson"))
## correlation sample wise
corr_metab_MTBE 

mean_corr_metab_MTBE <- mean(corr_metab_MTBE)
mean_corr_metab_MTBE

# metab and autoSP3
corr_metab_autoSP3 <- diag(cor(matrix_metab, matrix_autoSP3, method= "pearson"))
## correlation sample wise
corr_metab_autoSP3 

mean_corr_metab_autoSP3 <- mean(corr_metab_autoSP3)
mean_corr_metab_autoSP3
```

```{r}
t.test(corr_metab_autoSP3,corr_metab_MTBE)
```
# Visualize coor metab and prot1/2
```{r}
ph_corr_m <- matrix(NA, ncol = 2, nrow=5)

colnames(ph_corr_m) <- c("Metab and MTBE", "Metab and autoSP3")
rownames(ph_corr_m) <- c("Patient_A", "Patient_B", "Patient_C", "Patient_D", "Average")


ph_corr_m[c(1:4),1] <- corr_metab_MTBE
ph_corr_m[5,1] <- mean_corr_metab_MTBE

ph_corr_m[c(1:4),2] <- corr_metab_autoSP3
ph_corr_m[5,2] <- mean_corr_metab_autoSP3

breaks <- c(seq(0,0.99, by=.01))
cols <- c(colorRampPalette(c("white","#7285A5"))(length(breaks))) # or use length(breaks)

heatmap_plot_m <- ComplexHeatmap::pheatmap(ph_corr_m, display_numbers = T, cluster_cols = F,  cluster_rows=F, fontsize_number = 14, treeheight_row = 0, color = cols, breaks = breaks, column_names_side = c("top"), fontsize = 12, angle_col = c("0"), annotation_legend =T)
heatmap_plot_m

pdf("pHeatmap_pathway_enrichment_scores_corr_metab_and_proteomics_color.pdf", height = 3.5)
heatmap_plot_m
dev.off()

```

